Visualizing demographic evolution using geographically inconsistent census data

### 2019-08-10 paper/intro.tex
more fixes, regionalisation ideas

These advances have made neighbourhood constructs more tractable to quantitative
analysis. Yet new challenges have also emerged, especially at the convergence of
research on neighbourhood effects and neighbourhood dynamics. Neighbourhood
effects research assumes knowledge about the nature and scope of "the
neighbourhood" that presumably shapes individual
outcomes~\citep{Kwan2018,Shelton2019}. Concurrently, researchers note that
neighbourhoods are not necessarily fixed containers in which other processes
occur, but themselves dynamically
evolve~\citep{Delmelle2017,Reades2019,li2018new}. The result is to open up key
assumptions about neighbourhoods for theoretical and empirical examination: how
do we appropriately define and compare neighbourhoods at a given time?; how do
we appropriately define and compare the temporal trajectories of
neighbourhoods?; and can we do both at once, "fully
interactionally"~\citep{Abbott1997}: classify neighbourhoods now based on where
they came from and where they are going?

@@

Instead, we propose a network representation. Intuitively, a \emph{network}
(also called a \emph{graph}) is a collection of entities (nodes) that are
related to each other (edges). In this case, each different aggregation area is
represented as a node and we connect nodes that have overlapping geographical
areas, leading to the network illustrated on the right of
Figure~\ref{fig:intuition}. By partitioning the network into connected nodes
that are similar, we are effectively identifying clusters in the spatio-temporal
data, as illustrated by the colours of the nodes on the right side of
Figure~\ref{fig:intuition}. Further, all the possible paths of change can be
obtained by computing sequences of nodes over time, in this case: (A, C, F), (A,
D, F), (A, D, G), and (B, E, H). This representation is also suited for
geographically consistent regions, as illustrated by the stable park in this
example, and is therefore a generalisation of the traditional paradigm.

>>>>

\revision{In this work, we consider neighbourhoods as \emph{formal
regions}~\citep{montello2003regions}, geographically continous areas with
similar data characteristics.} These advances have made neighbourhood constructs
more tractable to quantitative analysis. Yet new challenges have also emerged,
especially at the convergence of research on neighbourhood effects and
neighbourhood dynamics. Neighbourhood effects research assumes knowledge about
the nature and scope of "the neighbourhood" that presumably shapes individual
outcomes~\citep{Kwan2018,Shelton2019}. Concurrently, researchers note that
neighbourhoods are not necessarily fixed containers in which other processes
occur, but themselves dynamically
evolve~\citep{Delmelle2017,Reades2019,li2018new}. The result is to open up key
assumptions about neighbourhoods for theoretical and empirical examination: how
do we appropriately define and compare neighbourhoods at a given time?; how do
we appropriately define and compare the temporal trajectories of
neighbourhoods?; and can we do both at once, "fully
interactionally"~\citep{Abbott1997}: classify neighbourhoods now based on where
they came from and where they are going?

@@

Instead, we propose a network-based representation. A \emph{network} (also
called a \emph{graph}) is a collection of entities (nodes) that are related to
each other (edges). In this case, each different aggregation area is represented
as a node and we connect nodes that have overlapping geographical areas\revision{ in
different times or are neighbours in the same time}, leading to the network
illustrated on the right of Figure~\ref{fig:intuition}. By partitioning the
network into connected nodes that are similar, we are effectively identifying
clusters in the spatio-temporal data, as illustrated by the colours of the nodes
on the right side of Figure~\ref{fig:intuition}. Further, all the possible paths
of change can be obtained by computing sequences of nodes over time, in this
case: (A, C, F), (A, D, F), (A, D, G), and (B, E, H). This representation is
also suited for geographically consistent regions, as illustrated by the stable
park in this example, and is therefore a generalisation of the traditional
paradigm.

### 2019-08-10 paper/related.tex
more fixes, regionalisation ideas

\subsection{Data clustering}

Data clustering is one of the elementary processes for data analysis,
simplifying the data into a smaller number of homogeneous sets that can be
interpreted in the same way. There is no shortage of contributions for this
problem~\citep{Fahad2014}, but most neighbourhood related applications still
rely on k-means~\citep{jain2010data,Delmelle2016} and, to a lesser extent, Self
Organising Maps~\citep{Delmelle2017,Ling2016}.

However, a method for geographic data analysis should not ignore the geographic
component of the data, and we extend research exploring ways to incorporate it.
One straightforward option for agglomerative methods~\citep{han2001spatial} is
to consider only nearby clusters for merging~\citep{Chavent2017}, which can also
be done for k-means~\citep{soor2018extending}. Alternatively, the spatial
distance could be directly added to the inter-cluster metric~\citep{Chavent2017}
via a mixing parameter, which adds flexibility to the method, but introduces the
problem of finding the correct application-dependent values.

Indeed, one crucial step in most clustering algorithms is the definition of the
number of clusters. We avoid this problem by considering hierarchical
methods~\citep{soille2012morphological}, where the result is not a partition of
the data, but a tree of partitions, similar to a dendogram. This approach is
interesting for interactive methods, because it allows the user to change the
number of clusters on the fly. Since our data is represented as a network, we
opted for an heuristic variation of the maximum weighted matching algorithm
called \emph{sorted maximal matching}~\citep{markus2017}, which merges clusters
based on the weights of the edges between pairs of clusters.

@@

Using data clustering, we can represent the region's cluster instead of all the
its variables~\citep{Alce2018,Valdivia2015,VonLandesberger2016}. While this
simplifies the geographic portion of the visualization, it introduces the
problem of how to summarise the contents of each cluster. One traditional
approach is to use parallel coordinates plot~\citep{ferreira2015urbane}, but
these they can get cluttered representing similar clusters over several
variables. Further, for demographic applications, the clusters are usually
strongly characterised by a small subset of
values~\citep{Delmelle2016,Delmelle2017}. Therefore, in the proposed method, we
identify the variables that are most relevant to the characterisation of each
cluster. The distribution of values on that variable is then represented using a
boxplot, a well known statistical plot displaying basic properties of the
distributions.

>>>>

\subsection{Data clustering and regionalisation}

Data clustering is one of the elementary processes for data analysis,
simplifying the data into a smaller number of homogeneous sets that can be
interpreted in the same way. There is no shortage of contributions for this
problem~\citep{Fahad2014}\revision{, since variations of it appear in almost all
scientific fields. Most neighbourhood related applications still rely on
k-means~\citep{jain2010data,Delmelle2016}, the Louvain method for community
detection~\citep{blondel2008Louvain,Thomas2012}, or, to a lesser extent, Self
Organising Maps~\citep{Delmelle2017,Ling2016,Dani2013_som_maxp}.}

\revision{In geography, this problem is known as
\emph{regionalisation}~\citep{montello2003regions}, a rather old problem that
has been throughly explored, leveraging different mathematical tools, including
discrete topology~\citep{brantingham1978topological} and discrete
geometry~\citep{assunccao2006efficient}. Indeed, network-based methods are among
the current state-of-the-art~\citep{Guo2008Redcap,maxp}. However,
\emph{temporal} regionalisation is significantly less explored, especially in a
demographic context, arguably due to the difficulties in dealing with
unharmonised longitudinal data.

Since we adopted a network-based data representation and our objectives include
an interactive interface, we opted for an heuristic variation of the maximum
weighted matching algorithm called \emph{sorted maximal
matching}~\citep{markus2017}, because of its simplicity, customisability, and
fast computation times. This algorithm merges clusters based on the weights of
the edges between pairs of clusters, creating an hierarchy over the data,
allowing the user to change the number of clusters without reprocessing the
data. Changing the clustering algorithm would lead to different results, but any
hierarchical network clustering method can be used in our framework. Indeed, our
core contribution is a different interpretation of the data representation, not
a new clustering or regionalisation method.}

@@

Using data clustering, we can represent the region's cluster instead of all the
its variables~\citep{Alce2018,Valdivia2015,VonLandesberger2016}. While this
simplifies the geographic portion of the visualisation, it introduces the
problem of how to summarise the contents of each cluster. One traditional
approach is to use parallel coordinates plot~\citep{ferreira2015urbane}, but
these they can get cluttered representing similar clusters over several
variables. Further, for demographic applications, the clusters are usually
strongly characterised by a small subset of
values~\citep{Delmelle2016,Delmelle2017}. Therefore, in the proposed method, we
identify the variables that are most relevant to the characterisation of each
cluster. The distribution of values on that variable is then represented using a
boxplot, a well known statistical plot displaying basic properties of the
distributions.

### 2019-08-10 paper/study.tex
more fixes, regionalisation ideas

\smallskip While Toronto is more dynamic than Los Angeles, possibly due to size
differences, the volatile regions shown in Figure~\ref{fig:toVol} did not change
as quickly or dramatically as the ones shown in Figure~\ref{fig:laVol}, which
involved twice as many people. We found this trend to be related to the
countries themselves, Canadian cities have larger areas undergoing slow, gradual
changes, whereas American cities have more general stability, but quicker
changes in smaller scales. The supplementary material contains brief summaries
of all the regions accessible in this prototype.

>>>>

% \smallskip While Toronto is more dynamic than Los Angeles, possibly due to size
% differences, the volatile regions shown in Figure~\ref{fig:toVol} did not change
% as quickly or dramatically as the ones shown in Figure~\ref{fig:laVol}, which
% involved twice as many people. We found this trend to be related to the
% countries themselves, Canadian cities have larger areas undergoing slow, gradual
% changes, whereas American cities have more general stability, but quicker
% changes in smaller scales. The supplementary material contains brief summaries
% of all the regions accessible in this prototype.
